Reproducible Research: Peer Assessment 1
======================================================

<br>

## Loading and preprocessing the data

```{r load libraries, message=FALSE}
library(ggplot2)                                # load libraries used in this assignment
library(dplyr)

data <- read.csv("activity.csv")                # read the data
data <- tbl_df(data)                            # convert to dplyr data frame
```

<br>

## What is mean total number of steps taken per day?

First, plot a histogram of the data

```{r histogram, fig.width = 10, message = FALSE}
dailySum <- group_by(data, date)                          # group data by date
dailySum <- summarize(dailySum, Steps = sum(steps))       # sum by date

h <- ggplot(dailySum, aes(x = Steps)) + 
        geom_histogram(fill = "skyblue", color = "darkgray") +
        labs(title = "Histogram of Total Daily Steps\n") +
        scale_y_discrete(breaks = seq(0, 10, by = 1))

print(h)
```

Next, calculate the Mean and Median of Total Daily Steps

```{r, mean and median}
sMean <- round(mean(dailySum$Steps, na.rm = T), 2)        # round to 2 decimal places
sMedian <- median(dailySum$Steps, na.rm = T)
```

#### Mean = `r format(sMean, nsmall=2)`  
#### Median = `r sMedian`

<br>

## What is the average daily activity pattern?

First, create a time series plot of the 5-minute intervals and the average number of steps
taken (averaged across all days)

```{r time series plot, fig.width = 10}
avgIntv <- group_by(data, interval)                             # group data by interval
avgIntv <- summarize(avgIntv, Steps = mean(steps, na.rm = T))   # average by interval

tsPlot <- ggplot(avgIntv, aes(x = interval, y = Steps)) + 
        geom_line(color = "blue") +
        labs(title = "Plot of Average Steps Taken per Interval (Across all Days)\n") +
        scale_x_discrete(breaks = seq(0, 2400, by = 100))

print(tsPlot)
```

Next, determine which 5-minute interval contains the maximum average number of steps

```{r interval with max steps}
maxSteps <- max(avgIntv$Steps)                          # finds max value in Steps column
maxIntv <- with(avgIntv, interval[Steps==maxSteps])     # finds which interval it is
```

#### Interval `r maxIntv` has the maximum average number of steps = `r round(maxSteps, 2)`

<br>

## Imputing missing values


<br>

## Are there differences in activity patterns between weekdays and weekends?
