Reproducible Research: Peer Assessment 1
======================================================

<br>

## Loading and preprocessing the data

```{r load libraries, message=FALSE}
library(ggplot2)                                # load libraries used in this assignment
library(dplyr)

data <- read.csv("activity.csv")                # read the data
data <- tbl_df(data)                            # convert to dplyr data frame
```

<br>



## What is mean total number of steps taken per day?

First, plot a histogram of the data

```{r histogram, fig.width = 10, message = FALSE}
dailySum <- group_by(data, date)                          # group data by date
dailySum <- summarize(dailySum, Steps = sum(steps))       # sum by date

h <- ggplot(dailySum, aes(x = Steps)) + 
        geom_histogram(fill = "skyblue", color = "darkgray") +
        labs(title = "Histogram of Total Daily Steps\n") +
        scale_y_discrete(breaks = seq(0, 10, by = 1))

print(h)
```

Next, calculate the Mean and Median of Total Daily Steps

```{r, mean and median}
sMean <- round(mean(dailySum$Steps, na.rm = T), 2)        # round to 2 decimal places
sMedian <- median(dailySum$Steps, na.rm = T)
```

#### Mean = `r format(sMean, nsmall=2)`  
#### Median = `r sMedian`

<br>



## What is the average daily activity pattern?

First, create a time series plot of the 5-minute intervals and the average number of steps
taken (averaged across all days)

```{r time series plot, fig.width = 10}
avgIntv <- group_by(data, interval)                             # group data by interval
avgIntv <- summarize(avgIntv, Steps = mean(steps, na.rm = T))   # average by interval

tsPlot <- ggplot(avgIntv, aes(x = interval, y = Steps)) + 
        geom_line(color = "blue") +
        labs(title = "Plot of Average Steps Taken per Interval (Across all Days)\n") +
        scale_x_discrete(breaks = seq(0, 2400, by = 100))

print(tsPlot)
```

Next, determine which 5-minute interval contains the maximum average number of steps

```{r interval with max steps}
maxSteps <- max(avgIntv$Steps)                          # finds max value in Steps column
maxIntv <- with(avgIntv, interval[Steps==maxSteps])     # finds which interval it is
```

#### Interval `r maxIntv` has the maximum average number of steps = `r round(maxSteps, 2)`

<br>



## Imputing missing values

First, determine the total number of missing values in the dataset (i.e. total number of
rows with NAs)

```{r}
totalNAs <- is.na(data$steps)           # creates a logical vector of which rows have NAs
totalNAs <- sum(as.numeric(totalNAs))   # convert TRUEs to 1 and sum
```

#### Total number of missing values is `r totalNAs`

Next, my strategy for filling in missing values in the dataset is to replace each NA value with the mean (calculated above) for that 5-minute interval it is found in.

```{r}
# convert the data back to a normal data frame
data2 <- data.frame(data)
# find the row numbers in the dataset that are NAs (i.e. missing)
rowNums <- as.integer(rownames(data2[is.na(data2$steps), ]))
# get the intervals the missing data belongs to
vIntervals <- data2[rowNums, "interval"]

# loop on each row that is missing data and replace it with the mean value corresponding
# to its 5-minute interval
for (i in 1:length(rowNums)) {
        data2[rowNums[i], "steps"] <- avgIntv[avgIntv$interval==vIntervals[i], "Steps"]
}
```

Now, with the missing values imputed, create a histogram of the of the total number of steps taken each day and then calculate the Mean and Median.

```{r, imputed histogram, fig.width = 10, message = FALSE}
dailySum2 <- group_by(data2, date)                          # group data by date
dailySum2 <- summarize(dailySum2, Steps = sum(steps))       # sum by date

h2 <- ggplot(dailySum2, aes(x = Steps)) + 
        geom_histogram(fill = "skyblue", color = "darkgray") +
        labs(title = "Histogram of Total Daily Steps (Imputed Data)\n") +
        scale_y_discrete(breaks = seq(0, 12, by = 1))

print(h2)

sMean2 <- mean(dailySum2$Steps)
sMedian2 <- median(dailySum2$Steps)
```

#### Mean = `r format(sMean2, nsmall=2)`  
#### Median = `r format(sMedian2, nsmall=2)`

It appears my strategy for imputing the data only increased the number of days the data set was near the mean, which the median almost proves. It went from 4 days observed in its bucket to 12.

<br>

## Are there differences in activity patterns between weekdays and weekends?

First, I'm going to add a new column that designates whether it is weekday or weekend and then average the imputed data in each 5-minute interval across the common day type. Then I'm going to plot two time series charts that compare weekday steps to weekend steps.

```{r panel plot, fig.width = 10}
data2 <- tbl_df(data2)

data2 <- mutate(data2, day = weekdays(as.Date(date)))
data2$day <- gsub("Saturday", "weekend", data2$day)
data2$day <- gsub("Sunday", "weekend", data2$day)
data2$day <- gsub("Monday", "weekday", data2$day)
data2$day <- gsub("Tuesday", "weekday", data2$day)
data2$day <- gsub("Wednesday", "weekday", data2$day)
data2$day <- gsub("Thursday", "weekday", data2$day)
data2$day <- gsub("Friday", "weekday", data2$day)

avgIntv2 <- group_by(data2, interval, day)              # group data by day by interval
avgIntv2 <- summarize(avgIntv2, Steps = mean(steps))    # average by interval

tsPlot2 <- ggplot(avgIntv2, aes(x = interval, y = Steps)) + 
        geom_line(color = "blue") +
        facet_wrap(~ day, ncol=1) +
        labs(title = "Plot of Average Steps Taken per Interval (Weekday vs. Weekend)\n") +
        scale_x_discrete(breaks = seq(0, 2400, by = 100))

print(tsPlot2)
```

On weekdays there appears to be more steps taken in the morning and evening, whereas weekends show more steps during the middle of the day. 


